---
title: "Simulation"
output: html_document
date: "2023-05-22"
---

```{r setup, include=FALSE}
library(PhaseTypeR)
library(tidyverse)
library(expm)
source('generate_subint_matrix.R')
```


## Method of moments: Simulation, n = 2, D = 2, M = 1
Estimating mutation rate (theta) and migration rate (M) from simulated data

Simulating data with theta = 5, M = 1:
```{r}
S <- generate_subint_matrix(n=2, D=2)$subint_mat #M is by default 1 in this case

theta <- 5 
nsim <- 1e4

T_mrca_w <- PH(S, init_probs = c(0.5,0,0.5))
T_mrca_b <- PH(S, init_probs = c(0, 1, 0))

T_total_w <- reward_phase_type(T_mrca_w, c(2,2,2))
T_total_b <- reward_phase_type(T_mrca_b, c(2,2,2))

sims_t_total_w <- rPH(nsim, T_total_w) 
sims_t_total_b <- rPH(nsim, T_total_b) 

mutations_w <- rpois(nsim, theta/2 * c(sims_t_total_w))
mutations_b <- rpois(nsim, theta/2 * c(sims_t_total_b))


emp_mean_w <- mean(mutations_w)
emp_var_w <- var(mutations_w) 


emp_mean_b <- mean(mutations_b)
emp_var_b <- var(mutations_b) 
```

### Estimation of migration rate and mutation rate using method of moments
For n = 2, D = 2, we can analytically handle the set of equations.

#### Within: 
results by setting sample means and variance equal to theoretical ones (found in Wakeley, 2009 p. 97)
and solving the two equations with two unknowns. 
```{r}
theta_hat_w_mat <- emp_mean_w/2
M_hat_w_mat <- -(emp_mean_w^2/(2*(emp_mean_w^2+emp_mean_w-emp_var_w)))

print(c(theta_hat_w_mat, M_hat_w_mat))
```
#### Between:
Same procedure, not so pretty formulas in this case:
```{r}
theta_hat_b_mat <- -(emp_mean_b*(-sqrt(-emp_mean_b^2*(3*emp_mean_b^2+4*emp_mean_b-4*emp_var_b)) + emp_mean_b^2 + 2*emp_mean_b - 2*emp_var_b )) / (2*(sqrt(-emp_mean_b^2*(3*emp_mean_b^2 + 4*emp_mean_b - 4*emp_var_b)) +emp_mean_b^2 ) )
theta_hat_b_mat

m_hat_b_mat <- (sqrt(-emp_mean_b^2*(3*emp_mean_b^2 + 4*emp_mean_b - 4*emp_var_b)) - emp_mean_b^2 - 2*emp_mean_b + 2*emp_var_b)/ (4*(emp_mean_b^2 + emp_mean_b - emp_var_b) )
m_hat_b_mat
```

### Numerical optimization with Method of moments method using optim.

For higher n and D we can use numerical optimization instead of solving the set of equations. 
We simulate data and let `optim` find the best suiting set of parameters by defining a loss function that compares theoretical moments using a set of parameters to the simulated moments.
```{r warning=FALSE}
set.seed(1)
n = 3
D = 3
within = F

theta1 <- 0.5
m1 <- 1

nsim <- 1e4

S1 <- generate_subint_matrix(n=n, D=D, M = matrix(rep(m1, D*D), nrow = D, ncol = D))$subint_mat 


if(within) {
  alpha <- c(1, rep(0, matrix_size(n,D)-1))
} else {
  alpha <- c(0, 1, rep(0, matrix_size(n,D)-2))
}

T_mrca <- PH(S1, init_probs = alpha)
T_total <- reward_phase_type(T_mrca, head(rowSums(state_space(n,D)), -1))

sims_t_total <- rPH(nsim, T_total) 

mutations <- rpois(nsim, theta1/2 * c(sims_t_total))

empirical_mean <- mean(mutations)
empirical_var <- var(mutations)

loss_function <- function(X) {
  theta = X[1]
  m = X[2]
  
  S <- generate_subint_matrix(n, D, M = matrix(rep(m, D*D), nrow = D, ncol = D))$subint_mat 
  reward_vector = head(rowSums(state_space(n,D)), -1)
  
  T_total_temp = reward_phase_type(PH(S, init_probs = alpha), reward = reward_vector)
  mean_t_total = mean(T_total_temp)
  var_t_total = var(T_total_temp)


  mean_S = (theta/2)*mean_t_total

  var_S = mean_S + (theta/2)^2 *var_t_total
  
  #compare with empirical mean and variance calculated from the previous simulations

  return(mean(abs(c(mean_S - empirical_mean, var_S - empirical_var))))
}


optim(c(1,1),loss_function, control = list(maxit = 10000))
```

## Maximum likelihood estimation

This method uses the sampling formula (Hobolth, 2023)

Helper functions:
```{r}
SFS_structured <- function(theta,n=3,D=2,within=T, m){
  if (within) {
    initVec <- c(1, rep(0,matrix_size(n,D)-1))
  }
  else {
    initVec <- c(0,1,rep(0,matrix_size(n,D)-2))
  }
  SMat <- generate_subint_matrix(n,D,M = matrix(rep(m, D*D), ncol=D, nrow=D))$subint_mat
  RMat <- c(head(rowSums(state_space(n,D)), -1))
  out <- list()
  out$initVec <- initVec
  out$SMat <- SMat
  out$RMat <- RMat
  out$mutaRate <- theta/2
  return(out)
}

evo_model_to_prob_model <- function(evo_model){
  SMat <- evo_model$SMat
  RMat <- evo_model$RMat
  lam <- evo_model$mutaRate
  p_total <- solve(diag(dim(SMat)[1])  - (1/lam)*(diag(1/RMat) %*% SMat))
  ## exit vector
  nSt <- nrow(SMat)
  sVec <- -SMat %*% rep(1,nSt)                 
  exitVec <- (diag(dim(SMat)[1]) - p_total) %*% matrix(rep(1, dim(SMat)[1]), nrow = dim(SMat)[1])
  out <- list()
  out$initVec <- evo_model$initVec
  out$exitVec <- exitVec
  out$ptotal <- p_total
  return(out)
}

#probability of observing xi = k total mutations. ´
#q and Ptotal supplied from `evo_model_to_prob_model` as exitVec and ptotal respectively.
prob_xi_total <- function(k,Ptotal,alpha,q){
  probs <- rep(NA, length(k))
  count = 1
  for (i in k) {
    probs[count] <- alpha %*% (Ptotal%^%i) %*% q
    count = count + 1
  }
  return(probs)
}

```

### Numerical optimization with MLE method using optim
```{r}
set.seed(1)
n = 3
D = 3
theta <- 0.5
within = F

nsim <- 1e4

S <- generate_subint_matrix(n, D, M = matrix(rep(1, D*D), nrow = D, ncol = D))$subint_mat 
reward_vector <- c(head(rowSums(state_space(n,D)), -1))

if (within) {
  alpha = c(1, rep(0, matrix_size(n,D)-1))
} else {
  alpha = c(0, 1, rep(0, matrix_size(n,D)-2))
}
T_total <- reward_phase_type(PH(S, init_probs = alpha), reward = reward_vector)

sims_t_total <- rPH(nsim, T_total)
mutations <- rpois(nsim, theta/2 * c(sims_t_total)) 

loss_mle <- function(X) {
  #theta = X[1]
  m = X[1]
  
  #find m and theta, which makes observing true k from sim most likely
  evo_model <- SFS_structured(theta=0.5,n=n,D=D,within=within, m=m)
  #Converting to probability model
  prob_model <- evo_model_to_prob_model(evo_model)

  #Calculating probability of observing the observed amount of mutations in the sample (total) (log-likelihood)
  prob_sample <- prob_xi_total(mutations,
                               prob_model$ptotal,
                               prob_model$initVec,
                               prob_model$exitVec) %>% log %>% sum
  return(-prob_sample)
}

#L-BFGS-B allows bounds on the search space
optim(0.1, loss_mle, method = 'L-BFGS-B', lower = 0.1, upper = 5)
```
within:  0.4927916
between: 0.4965742


## Singletons and doubletons branch length

Here, we sample singleton and doubleton branch length for n = 3, D = 2, instead of sampling the total number of mutations. 

Helper functions (almost equivalent to above ones):
```{r}
SFS_Three_structured_branches <- function(theta,n=3,D=2,within=T, m){
  if (within) {
    initVec <- c(1, rep(0,matrix_size(n,D)-1))
  }
  else {
    initVec <- c(0,1,rep(0,matrix_size(n,D)-2))
  }
  SMat <- generate_subint_matrix(n,D,M = matrix(rep(m, D*D), ncol=D, nrow=D))$subint_mat
  RMat <- matrix( c(3,0, 3,0, 3,0, 3,0, 1,1, 1,1, 1,1), ncol=2, byrow=TRUE)
  out <- list()
  out$initVec <- initVec
  out$SMat <- SMat
  out$RMat <- RMat
  out$mutaRate <- theta/2
  return(out)
}

evo_model_to_prob_model_2 <- function(evo_model){
  SMat <- evo_model$SMat
  RMat <- evo_model$RMat
  lam <- evo_model$mutaRate
  ## A matrix
  AMat <- solve( -SMat+lam*( diag( rowSums( RMat ) ) ) )
  ## exit vector
  nSt <- nrow( SMat )
  sVec <- -SMat %*% rep(1,nSt)                 
  exitVec <- AMat %*% sVec
  ## Sub-transition matrices
  prob_array <- array( 0, dim = c( ncol(RMat), rep(nSt, 2) ) )
  for (i in 1:ncol(RMat)){
    prob_array[i,,] <- AMat %*% ( lam*diag( RMat[,i] )) 
  }
  out <- list()
  out$initVec <- evo_model$initVec
  out$prob_array <- prob_array
  out$exitVec <- exitVec
  return(out)
}


sample_prob <- function(sample,prob_model){
  initVec <- prob_model$initVec
  prob_array <- prob_model$prob_array
  exitVec <- prob_model$exitVec
  ## Zero mutations
  if (sum(sample)==0) return( as.double( initVec %*% exitVec) )
  ## Positive number of mutations
  nSt <- length(initVec)
  mSet <- multiset( c( rep( 1:length(sample), sample ) ) )  
  sum_prb <- 0
  for (k in 1:ncol(mSet)){
    ProdMat <- diag(nSt)
    for (m in 1:nrow(mSet)){
      ProdMat <-  ProdMat %*% prob_array[mSet[m,k],,]
    }
    sum_prb <- sum_prb + initVec %*% ProdMat %*% exitVec
  }
  return( as.double( sum_prb ) )
}

#vectorized version: can take list of pairs, e.g. 'samples <- list(c(2,3), c(1,0), c(5,4))'
vec_sample_prob <- Vectorize(sample_prob, 'sample')
```


### Simulating singleton and doubleton mutations
```{r}
set.seed(1)

theta <- 0.5
#Evolutionary model
sfs_model_branches <- SFS_Three_structured_branches(theta=theta, m = 1, within = T)
#Converting to probability model
prob_model <- evo_model_to_prob_model_2(sfs_model_branches)

#sample_prob(c(2,1), PrbMdl = prob_model)

#multivariate phase type of singleton and doubleton branch lengths
single_double_lengths <- MPH(sfs_model_branches$SMat, reward_mat = matrix( c(3,0, 3,0, 3,0, 3,0, 1,1, 1,1, 1,1), ncol=2, byrow=TRUE))


nsim <- 1e4

length_sims_single_double <- rMPH(nsim, single_double_lengths)

sims_singleton_mut <- rpois(nsim, lambda = length_sims_single_double[,1]*theta/2)
sims_doubleton_mut <- rpois(nsim, lambda = length_sims_single_double[,2]*theta/2)

list_samples <- Map(c, sims_singleton_mut, sims_doubleton_mut)
```

### Numerical optimization 
```{r}
loss_branches <- function(X) {
  TH = X[1]
  M = X[2]
  
  sfs_model_branches <- SFS_Three_structured_branches(theta = TH, m = M, within = T)
  
  
  #Converting to probability model
  prob_model <- evo_model_to_prob_model_2(sfs_model_branches)
  
  
  prob_sample <- vec_sample_prob(list_samples, prob_model) %>% log %>% sum
  return(-prob_sample)
}


optim(c(1,3), loss_branches, method = 'L-BFGS-B', lower = c(0.1, 0.1), upper = c(10,10))
```





